---
title: "Cumulative Accessibility Measure in R"
output: rmarkdown::md_document
---
  
To conduct an accessibility analysis, we need:  
1) OSM Road Data  
2) Shapefile of the study area  (with population and employment distribution)  
3) GTFS feed   

We use these inputs to make queries on Open Trip Planner

The shapefile and GTFS feed have been made available by Transport for Cairo, an Egyptian public transport consultancy

### Part 1: OSM Data

Define the geographic extents as a vector of c(xmin, ymin, xmax, ymax)



You can plot to see if the roads have been downloaded. This takes around 10 minutes and the sp file is huge.
I will not do it here, but for checking purposes, this is how it would be done:

1) convert xml to sp using osmdata_sp(query)

2) plot sp::plot(highways_greaterCairo$osm_lines)
highways_greaterCairo <- osmdata_sp(query)


An alternative to xml is pbf files. These are more compact and presumably make the forthcoming analysis faster.
They can be downloaded from the HOT OSM export tool https://export.hotosm.org/en/v3/exports/new/formats?fbclid=IwAR3m_1bDZK2sYsA-jtPRPYGg9R7kTqt5hzjP88x3p2yvRd4i9tr11i2tG60

I tried to use a pbf file but the analysis was not quicker so I stuck to the xml file extracted from r

### Part 2: Importing Shapefiles for Analysis

In this step I import a shapefile of Greater Cairo. The shapefile has the region divided into hexagons with the population and number of jobs in each hexagon

```{r}
library(sf)

# import the shapefile
cairo_hexagons <- st_read("Cairo Shapefiles/H3-res8/H3res-8_GCR_4326.shp")
```

It is imported and in the corrected crs EPSG:4326

Let's plot to check if it looks right  
```{r warning=FALSE}
library(tmap)

# plot
tm_shape(cairo_hexagons) +
              tm_polygons()
```

